Drought indices of water demand and supply, soil moisture, vegetation, and their impact on LULCC in continental Chile

### 2024-01-24 article/paper_drought_Chile2023.qmd
actualizar discusion

In this research, we aim to analyze the impact of drought on different types of ecosystems (land cover classes) in continental Chile. Our specific goals are: i) to analyze the trend on multi-scalar drought indices for water demand and supply, soil moisture, and vegetation productivity for 1981–2023/2001–2023; ii) to assess the LULC change for 2001–2021 and how it relates to drought indices; iii) to evaluate the relationship between zcNDVI, a measure of vegetation productivity, and drought indices for water demand and supply and soil moisture; and iv) to assess if the observed changes in the drought indices are linked to TWS.

@@

### land cover macroclasess and validation

@@

### land cover persistence mask 2001-2022

@@

### land cover trend and drought indices

@@

### land cover persistence mask 2001-2022

@@

Regarding vegetation productivity aggregated through the macrozones in the five land cover macroclasses, in "Norte Grande," there is an increase trend of 0.02 (z-index) per decade, related to types of grassland and shrubland. There is a negative trend in "Norte Chico" with -0.04 and "Centro" with -0.02 per decade. In the "Norte Chico," savanna (-0.05) has the lowest trend, and the rest of the types are around -0.04. In "Centro," shrubland reaches -0.06, grassland -0.05, and croplands and savanna -0.01 per decade. This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023]. Vegetation reached its lowest values since the year 2019, reaching an extreme condition in early 2020 and 2022 in the "Norte Chico" and Centro" (Mega Drought). The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var). Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the Chilean matorral appears to be the region most affected by a negative trend in vegetation [@Fuentes2021].

@@

## Drought trend, LULCC, and climate conditions

1.-  Respecto a lo que indica @Vicente-Serrano2018, de que el aumento en la tendencia en severidad de la sequía (hidrológica) tiene que ver más con un aumento de la demanda de agua (ej, LULCC, amazonas) que a una tendencia en las condiciones climáticas (SPI-12). ¿Qué pasa en Chile?

## land cover types most impacted by drought throughout Chile

@@

>>>>

In this research, we aim to analyze the impact of drought on different types of ecosystems (land cover classes) in continental Chile. Our specific goals are: i) to analyze the trend on multi-scalar drought indices for water demand and supply, soil moisture, and vegetation productivity for 1981–2023/2001–2023; ii) to assess the LULC change for 2001–2022 and how it relates to drought indices; iii) to evaluate the relationship between zcNDVI, a measure of vegetation productivity, and drought indices for water demand and supply and soil moisture; and iv) to assess if the observed changes in the drought indices are linked to TWS.

@@

### Land cover macroclasess and validation

@@

### Land cover persistence mask 2001-2022

@@

### Land cover trend and drought indices

@@

### Land cover persistence mask 2001-2022 {#sec-persistence_mask}

@@

Regarding vegetation productivity aggregated through the macrozones in the five land cover macroclasses, in "Norte Grande," there is an increase trend of 0.02 (z-index) per decade, related to types of grassland and shrubland. There is a negative trend in "Norte Chico" with -0.04 and "Centro" with -0.02 per decade. In the "Norte Chico," savanna (-0.05) has the lowest trend, and the rest of the types are around -0.04. In "Centro," shrubland reaches -0.06, grassland -0.05, and croplands and savanna -0.01 per decade. This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023]. Vegetation reached its lowest values since the year 2019, reaching an extreme condition in early 2020 and 2022 in the "Norte Chico" and Centro" (Mega Drought). The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var). Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the Chilean matorral [@Fuentes2021] appears to be the region most affected by a negative trend in vegetation productivity.

@@

## Drought trend and attribution to LULCC

@Vicente-Serrano2021, in a study at the global scale of drought trends, indicates that there have not been significant trends in meteorological drought since 1950. Also, state that the increase in hidrological trend in some parts of the globe (northeast Brazil and the Mediterranean region) is related to changes in land cover and specifically to the rapidly increasing irrigated area, which consequently increases water extraction. @Kogan2020 analyzed the agricultural drought impact globally and in the main grain producer countries, finding that _"since 1980, the Earth warming has not changed the drought area or intensity"_. In our study, we considered the variation in vegetation productivity in Chile for areas without changes in land cover macroclasses (see @sec-persistence_mask), to avoid misleading conclusions that could be related to the increase in water demand due to LULCC. Our results show a contrasting perspective. There has been a significant trend in the decline of vegetation productivity (zcNDVI) since 2000 for "Norte Chico" and "Centro," which has been extreme between 2020 and 2022, seemsly due to an intense hydrological drought due to the persistance of the Mega Drought [@Garreaud2017]. Despite we used the persistance mask for vegetations's trend analysis, the cropland, which is the most water-demand type, showed a decrease trend in "Norte Chico" and "Centro." Also, there was an increase in barren land for both types. These changes are associated with a decrease in water demand from vegetation. On the other hand, for "Norte Chico" and "Centro," our results show a decrease in trends of water supply (SPI and SSI), which are higher at larger time scales and consequently impact the hydrological system. We claim that what occurred in central Chile defies findings made at the global level [@Vicente-Serrano2021;@Kogan2020], demonstrating that a constant decrease in water supply rather than an increase in water demand (i.e., irrigated crops) is the main cause of the hydrological drought. Finally, central Chile has shown a diminishment in vegetation productivity for all macroclasses, mainly attributed to variation in water supply, i.e., precipitation, which could be strengthened by an increase in water demand by, for example, irrigated crops.

1.-  Respecto a lo que indica @Vicente-Serrano2022, de que el aumento en la tendencia en severidad de la sequía (hidrológica) tiene que ver más con un aumento de la demanda de agua (ej, LULCC, amazonas) que a una tendencia en las condiciones climáticas (SPI-12). ¿Qué pasa en Chile?

@Chen2022: Assessing the impact of drought-land cover change on global vegetation greenness and productivity

@Kogan2020: muestra que la sequía no se ha intensificado ni expandido a nivel global, por 38 años.

## Land cover types and their impact by drought

Extreme drought conditions are an important driver of tree mortality, as shown by @Senf2020 in Europe. However, we found that forest is the type of landcover macroclass less affected by variation in drought indices, being the most resistant landcover class to drought. Supporting this is @Fathi-Taperasht2022, who asserts that Indian forests are the most drought-resistant and recover rapidly. Similarly, the work of @Wu2024 who analyzed vegetation loss and recovery in response to meteorological drought in the humid subtropical Pearl River basin in China, indicates that forests showed higher drought resistance. @Xiao2023 using Vegetation Optical Depth (VOD), kNDVI, and EVI, evaluates ecosystem resistance, showing that, compared to croplands, ecosystems with a higher proportion of forests are more resilient to severe droughts. They attribute the difference to a deeper rooting depth of trees, a higher water storage capacity, and different water use strategies between them [@Xiao2023].

@@

@Fathi-Taperasht2022: muestra que bosques son más resilentes a la sequía en India.

@Wu2024: evalua la perdida de vegetación en diferentes tipos de uso de suelo. Muestra que bosques son más resilentes y cultivos y pastizales, los menos.

@Jiang2020: impacto de la sequía en landcover en China.
